### 2018-05-24 write-up/abstract.tex
Start Evaluation

\newpage
{\normalfont\Huge\sffamily\bfseries Abstract}
\vspace{24pt}

This is the abstract. Write a summary of the whole thing. Make 
sure it fits in one page.

>>>>

\newpage
{\normalfont\Huge\sffamily\bfseries TO DO: Abstract}
\vspace{24pt}

In this thesis, I argue that linear types are an appropriate, \emph{type-based
formalism} for expressing aliasing, read/write permissions, memory allocation,
re-use and deallocation, first, in the context of the APIs of linear algebra
libraries and then in the context of matrix-expression compilation.

I show that framing the problem using linear types can help \emph{reduce bugs}
by making precise and explicit the informal, ad-hoc practices typically
employed by human experts and linear algebra \emph{compilers} and automate
checking them.

As evidence for this argument, I show non-trivial, yet readable, linear algebra
programs, that are safe (with respect to aliasing, read/write permissions,
memory re-use and deallocation) and explicit (with respect to memory
allocation) which (1) are more memory-efficient than equivalent programs
written using purely high-level linear algebra libraries and (2) perform just
as predictably as equivalent programs written using low-level linear-algebra
libraries. I also argue \emph{the experience} of writing such programs with
linear types is ``better'' in key-respects. In addition to all of this, I show
that it is possible to provide such features \emph{as a library} on top of
existing programming languages and linear algebra libraries.

\textbf{Summarise evaluation.}

\textbf{Summary future work.}

### 2018-05-24 write-up/background.tex
Start Evaluation

\prechapter{%
    In this chapter, I will outline the concept of linear types and show how
    they can be used to solve the problems faced by programmers writing code
    using linear-algebra libraries. I will be emphasising the \emph{practical}
    and intuitive explanations of linear types to keep this thesis accessible
    to working programmers as well as academics not familiar with type-theory;
    giving only a terse overview of the history and theory behind linear types
    for the interested reader to pursue further.
}%

>>>>

\prechapter{%
    I will outline the concept of linear types and show how they can be used to
    solve the problems faced by programmers writing code using linear-algebra
    libraries. I will emphasise \emph{practical} and intuitive explanations of
    linear types to keep this thesis accessible to working programmers as well
    as academics not familiar with type-theory and will give only a terse
    overview of the history and theory behind linear types for the interested
    reader to pursue further.
}%

### 2018-05-24 write-up/evaluation.tex
Start implementation, related work and conclusion
Finish Background, continue Implementation
Start Evaluation

@@

>>>>

% Least Squares Linear Regression
% p73 onwards https://people.cs.uchicago.edu/~mrocklin/storage/dissertation.pdf

@@

\prechapter{%
    I will evaluate the central premise of this thesis: linear types are a
    practical and usable tool to help working programmers write readable, safe
    (with respect to aliasing, read/write permissions, memory re-use and
    deallocation) and explicit (with respect to memory allocation) code that
    (1) is more memory-efficient than code written using purely high-level
    linear algebra libraries and (2) performs just as predictably as code
    written using low-level linear-algebra libraries. I will also elaborate on
    the qualitative benefits of using linearly types to write linear-algebra
    programs.
}%

\section{Set-up}

I analysed four implmentations of a Kalman filter, increasing
the abstraction level at each step:

\begin{enumerate}

    \item a CBLAS/LAPACKE implementation, handwritten in C
        (Figure~\ref{fig:cblas_kalman}), with a minimal number of temporaries,
        calls to `symm' for matrices known to be symmetric ahead of time,
        transposition passed in to `gemm' as a flag and Cholesky decomposition
        for multiplying by an inverse of a matrix

    \item a LT4LA implementation (Figure~\ref{fig:ltfla_kalman}), also with a
        minimal number of temporaries, calls to `symm' for matrices known to be
        symmetric ahead of time, transposition passed in to `gemm' as a flag
        and Cholesky decomposition for multiplying by an inverse of a matrix

    \item an Owl/OCaml implementation using a Cholesky decomposition
        (Figure~\ref{fig:chol_owl_kalman}) but not taking advantage of matrices
        known to be symmetric ahead of time, and producing a new temporary
        matrix for every operation (including inverse and tranpose)

    \item an idiomatic Owl/OCaml implmentation
        (Figure~\ref{fig:chol_owl_kalman}) with an explicit inverse (LU
        decomposition), not taking advantage of matrices known to be symmetric
        ahead of time, and producing a new temporary matrix for every operation
        (including inverse and tranpose).

\end{enumerate}

\begin{sidewaysfigure}

    \begin{minted}[linenos, fontsize=\small]{c}
static void kalman(
    const int n,               const int k,                const double *sigma, /* n,n */
    const double *h, /* k,n */ const double *mu, /* n,1 */ double *r,           /* k,k */
    double *data,    /* k,1 */ double **ret_mu,  /* k,1 */ double **ret_sigma   /* n,n */
) {
        double* k_by_n = (double *) malloc(k * n * sizeof(double));
/*20*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, k, n, 1., sigma, n, h, n, 0., k_by_n, n);
/*21*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, k, k, n, 1., k_by_n, n, h, n, 1., r, k);
/*22*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
/*23*/  cblas_dcopy(k * n, h, 1, k_by_n, 1);
        double* k_by_k = (double *) malloc(k * k * sizeof(double));
/*24*/  cblas_dcopy(k * k, r, 1, k_by_k, 1);
/*25*/  LAPACKE_dposv(LAPACK_ROW_MAJOR, 'U', k, n, k_by_k, k, k_by_n, n);
/*27*/  LAPACKE_dpotrs(LAPACK_ROW_MAJOR, 'U', k, 1, k_by_k, k, data, 1);
        free(k_by_k);
        double* n_by_n = (double *) malloc(n * n * sizeof(double));
/*28*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, n, k, 1., h, n, k_by_n, n, 0., n_by_n, n);
        free(k_by_n);
        double* n_by_1 = (double *) malloc(n * sizeof(double));
/*29*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, 1, k, 1., h, n, data, 1, 0., n_by_1, 1);
        double* new_mu = (double *) malloc(n * sizeof(double));
/*30*/  cblas_dcopy(n, mu, 1, new_mu, 1);
/*31*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, 1, 1., sigma, n, n_by_1, 1, 1., new_mu, 1);
        free(n_by_1);
        double* n_by_n2 = (double *) malloc(n * n * sizeof(double));
/*32*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, n, n, 1., sigma, n, n_by_n, n, 0., n_by_n2, n);
/*33*/  cblas_dcopy(n*n, sigma, 1, n_by_n, 1);
/*34*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, n, -1., sigma, n, n_by_n2, n, 1., n_by_n, n);
        free(n_by_n2);
        *ret_sigma = n_by_n;
        *ret_mu = new_mu;
}
    \end{minted}
    \caption{CBLAS/LAPACKE implementation of a Kalman filter. I used C instead
        of Fortran because it is what Owl uses under the hood and OCaml FFI
        support for C is better and easier to use than that for Fortran. A distinct
        `measure\_kalman' function that sandwiches a call to this funtion with
        \texttt{getrusage} is omitted for brevity.}\label{fig:cblas_kalman}

\end{sidewaysfigure}

\begin{figure}[tp]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let potrs ~uplo a b =
  let b = Owl.Mat.copy b in
  Owl.Lapacke.potrs ~uplo ~a ~b
;;

let chol_kalman ~sigma ~h ~mu ~r ~data =
  let open Owl.Mat in
  let ( * ) = dot in
  let h' = transpose h in
  let sigma_h' = sigma * h' in
  let chol = Owl.Linalg.D.chol (r + h * sigma_h') in
  let sigma_h'_inv rest = sigma_h' * potrs ~uplo:'U' chol rest in
  let new_sigma = sigma - sigma_h'_inv (h * sigma) in
  let new_mu = mu + sigma_h'_inv (h * mu - data) in
  ((sigma, (h, (mu, (r, data)))), (new_mu, new_sigma))
;;

let owl_kalman ~sigma ~h ~mu ~r ~data =
  let open Owl.Mat in
  let ( * ) = dot in
  let h' = transpose h in
  let sigma_h' = sigma * h' in
  let x = sigma_h' * (inv @@ r + h * sigma_h') in
  let new_mu = mu + x * (h * mu - data) in
  let new_sigma = sigma - x * h * sigma in
  ((sigma, (h, (mu, (r, data)))), (new_mu, new_sigma))
;;
    \end{minted}
    \caption{Implementations of a Kalman filter using Owl, top one using a
        Cholesky decomposition, bottom one using idiomatic Owl. Owl does not
        yet provide a non-mutating `potrs' function, so I wrote my own which
        returns a mutated copy of its argument instead.}\label{fig:chol_owl_kalman}

\end{figure}

I evaluated the implmentations on two metrics: memory usage (via number and
size of temporaries allocated) and execution time. For the former, I compiled
Owl with print-statements on the relevant primitives to see exactly the number
of temporaries allocated for a single call of each function. While I did also
attempt to use gperftools and OCaml's profiling support with gprof for a more
holistic view of memory usage in the presence of OCaml's garbage-collector, I
ran into technical difficulties irrelevant to this thesis.

I measured execution time, in micro-seconds, against an exponentially (powers
of 5) increasing scaling factor for matrix size parameters $n=5$ and $k=3$.
For small scaling factors (1, 5, 25), I used the Core\_bench micro-benchmarking
library, for larger factors (125 and greater), I used the \texttt{getrusage}
system call (called \ltfla{Unix.times} in OCaml).  Core\_bench performs a
linear-regression (here, time against batch-size) so includes 95\%
confidence-interval with $R^2$. Larger scales have errors reported to $\pm
\sigma$ (one standard-deviation)

\section{Results}

Raw output of implementation traces and the benchmarking program (including
sample sizes) for each of the implementations are in
Appendix~\ref{chap:eval_data}.

\subsection{Memory Usage}

Inspecting the Owl trace shows generated 11 temporary matrices (13 calls to
empty, 2 of which are the resulting matrices). Chol shows 13 temporaries (same
as Chol plus two temporaries for the two calls to potrs), whereas both LT4LA
and CBLAS have 4 temporaries. Analysing the sub-expressions of the Owl
implementation shows the total amount of memory allocated for temporaries is $n
+ n^2 + 4nk + 3k^2 + 2k$ words; for Chol the total is that of Owl plus $n +
nk$; for LT4LA and CBLAS the total is $n + n^2 + nk + k^2$.

\subsection{Execution Time}

A graph of the execution times (with error bars which are present but quite
small) is show in Figure~\ref{fig:timings}.

\begin{figure}[tp]
    \centering
    \input{timings.tex}
    \caption{Comparison of execution times. Small matrices and timings $n \le
        5^3$ were micro-benchmarked with the Core\_bench library. Larger ones used
    Unix's \texttt{getrusage} functionality.}\label{fig:timings}
\end{figure}

For $n=5$, and a per implementation sample size of around 450, the CBLAS
implementation is the fastest ($24\mu s$), followed by LT4LA ($41 \mu s$), then
Cholesky and Owl (around $52 \mu s$).  However, for $n=25$, and around 350
measurements per implementation, the idiomatic Owl implementation is the fastest
($95 \mu s$), \emph{then} CBLAS ($104 \mu s$), with LT4LA and Chol last (around
$130 \mu s$). The 95\% confidence-intervals around these measurements have been
between 0 and 2.

For $n=125$, and a per implementation sample size of around 110, the trends change:
CBLAS implmentation is now the \emph{slowest} ($1803 \mu s\, [1746,\, 1867]$).
This is followed by LT4LA ($1678 \mu s\, [1646,\, 1714]$) and Owl ($1488 \mu
s\, [1464,\, 1515]$). Chol had a low $R^2$ value ($0.74$) and so its execution
time is inconclusive here.

For $n=625$ and a sample size of 1000, Chol was now the fastest ($125.5 \pm
25\,ms$), then Owl ($146 \pm 32 \,ms$), followed by LT4LA ($180.5 \pm 38 \,ms$)
and CBLAS ($188 \pm 36 \, ms$). Despite the large sample size, the
standard-deviation is still quite high; however, calculating the pair-wise
$p$-values (Welch's t-test) suggests that the difference in the means is meaningful.

For $n=3125$, Chol takes approximately half the execution time ($11.4 \pm 2.7
\,s$) of the rest (differences between LT4LA, Owl and CBLAS were not
statistically significant, $p>.05$, using Welch's t-test) around $22s$.

\section{TO DO: Analysis}

Having access to primitives which allow a programmer to re-use memory means
that memory usage for temporaries in LT4LA is on par with that of C. One caveat
is that the `free\_mat' primitive is a no-op in LT4LA, which still relies on
OCaml's garbage-collector. The difference between these two implementations
and the idiomatic Owl implementation is $k(3n+2k+2)$ words, and

\section{TO DO: Limitations}

\section{TO DO: Qualitative Benefits}

### 2018-05-24 write-up/implementation.tex
Start Evaluation

\prechapter{%
    I implemented LT4LA in OCaml, however I strongly believe the ideas
    described in this chapter can be applied easily to other languages and also
    are modular enough to extend the OCaml implementation to output to different
    back-end languages. I will show how a small core language with a few features
    can be the target of heavy-desugaring of typical linear-algebra programs. This
    core language can then be elaborated and checked for linearity before
    performing some simple and predictable optimisations and emitting (in this
    particular implementation) OCaml code that is not obviously safe and correct
    (with respect to linearity).
}%

@@

Code generation is a thankfully straightforward mapping from core LT4LA
constructs to OCaml constructs, with the appropriate sprinkling of \ltfla{Many}
constructors for integer, element and boolean literals.

@@

Even though Figure~\ref{fig:ocaml_kalman} is in OCaml, I see little reason that
it cannot be C or Fortran. Translating LT4LA's core language constructs to
those languages would not be as easy as translating to OCaml. However, it is
easy to see that both Figures~\ref{fig:fortran_kalman}
and~\ref{fig:ocaml_kalman} are realisations of a concise, linearly-typed Kalman
filter~\ref{fig:ltfla_kalman} \emph{specification} that describes the whole
behaviour and intent (with respect to read/writes permissions, memory
management and aliasing) of the program and the BLAS primitives it uses more
accurately than a Fortran, C or OCaml implementation could.

@@

So once you have written your memory-optimised program with all the features
and support provided by LT4LA and made it produce well-typed, compilable OCaml
code, the question then becomes, how to use this code. This process will vary
across ecosystems, but within OCaml, the new build-system on the block
\emph{Dune} has support for generating, compiling and linking OCaml modules at
\emph{compile time}. Indeed, this is how I have written tests and benchmarks
for programs produced by LT4LA from within OCaml. In particular, I can use my
linearly-typed Kalman filter implementation just like and with any other OCaml
function.

@@

I have explained how a few core features -- linearity, the \ltfla{Many}
constructor, value-restriction, fractional-capabilities with inference,
if-expressions and recursive functions -- are enough to \emph{statically
capture and automatically check} aliasing, read/write permissions, memory
allocation, re-use and deallocation of real world linear algebra library
routines. I have also demonstrated that simple pattern-matching and desugaring
provides the potential for a new, \emph{type-directed} approach to matrix
expression compilation. Lastly I have shown that it is possible to use these
features with \emph{existing} languages and frameworks.\footnote{As mentioned
in the previous chapter, if the host language supports
\emph{syntax-extensions}, like PPX for OCaml, it would be possible to use LT4LA
features \emph{from within} OCaml instead of just alongside it.}

>>>>

\prechapter{%
    I will describe the structure of LT4LA and explain the features of its core
    language. I will then show how typical linear-algebra programs can be
    elaborated into the core language and checked for linearity. Finally,
    I will explain how such programs can be translated (in this particular
    implementation) to OCaml code that is not obviously safe (with
    respect to aliasing, read/write permissions and memory management).
    Although I implemented LT4LA in OCaml, I believe the ideas described in
    this chapter can be applied easily to other languages and also are modular
    enough to extend the OCaml implementation to output to different back-end
    languages.
}%

@@

Code generation is a straightforward mapping from core LT4LA constructs to
OCaml constructs, with the addition of \ltfla{Many} constructors to wrap
integer, element and boolean literals.

@@

It is clear that both Figures~\ref{fig:fortran_kalman}
and~\ref{fig:ocaml_kalman} are realisations of a concise, linearly-typed Kalman
filter~\ref{fig:ltfla_kalman} \emph{specification} that describes the whole
behaviour and intent (with respect to read/writes permissions, memory
management and aliasing) of the program and the BLAS primitives it uses more
accurately than a Fortran, C or OCaml implementation could.

@@

So once you have written your memory-optimised program with all the features
and support provided by LT4LA and made it produce well-typed, compilable OCaml
code, the question then becomes, how to use this code. This process will vary
across ecosystems, but within OCaml, the new build-system on the block
\emph{Dune} has support for generating, compiling and linking OCaml modules at
\emph{compile time} (Figure~\ref{fig:build}). Indeed, this is how I have written
tests and benchmarks for programs produced by LT4LA from within OCaml. In
particular, I can use my linearly-typed Kalman filter implementation just like
and with any other OCaml function.

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{impl_build}
    \caption{Using LT4LA functions from OCaml with the Dune
        build-system.}\label{fig:build}
\end{figure}

@@

I have explained how a few core features -- linearity, the \ltfla{Many}
constructor, value-restriction, fractional-capabilities with inference,
if-expressions and recursive functions -- are enough to \emph{statically
capture and automatically check} aliasing, read/write permissions, memory
allocation, re-use and deallocation of non-trivial linear algebra programs. I
have also demonstrated that simple pattern-matching and desugaring provides the
potential for a new, \emph{type-directed} approach to matrix expression
compilation. Lastly I have shown that it is possible to use these features with
\emph{existing} languages and frameworks.\footnote{As mentioned in the previous
    chapter, if the host language supports \emph{syntax-extensions}, like PPX
for OCaml, it is possible to construct LT4LA expressions \emph{from within} the
host language.}
